
---
title: "World Happiness Report"
subtitle: "Happiness Trends and Contributing Factors (2011-2024)"
format: 
  html:
    code-fold: false
    code-summary: "Show code"
    embed-resources: true
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(patchwork)


whr <- read_csv(here::here("data/whr.csv"))
```

## Introduction

This report analyzes the data from the [2025 World Happiness Report](https://www.worldhappiness.report/), which looks at global happiness trends and factors that contribute to national happiness levels from 2011 to 2024. The dataset contains happiness rankings and scores for 169 countries.

## Global Happiness Trends

### Overall Trends (2011-2024)

```{r global-trends}
# Calculate global happiness statistics by year
global_trends <- 
  whr |>
  group_by(year) |>
  summarise(
    mean_happiness = mean(ladder_score, na.rm = TRUE),
    median_happiness = median(ladder_score, na.rm = TRUE),
    min_happiness = min(ladder_score, na.rm = TRUE),
    max_happiness = max(ladder_score, na.rm = TRUE),
    countries_count = n(),
    .groups = 'drop'
  )

global_trends |> 
  ggplot(aes(x = year, y = mean_happiness)) +
  geom_line(color = "steelblue", linewidth = 1.5) +
  geom_point(color = "steelblue", size = 3) +
  labs(
    title = "Global Average Happiness Over Time",
    subtitle = "World Happiness Report Data (2011-2024)",
    x = "Year",
    y = "Average Happiness Score"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  ) +
  scale_x_continuous(breaks = seq(2011, 2024, 1)) +
  scale_y_continuous(limits = c(5.2, 5.7), breaks = seq(5.2, 5.7, 0.1))
```

#### Key findings

- **Gradual Improvement**: Average global happiness has increased from ~5.39 (2011) to ~5.58 (2024)
- **2011-2016**: Relatively stable around 5.35-5.40
- **2017-2019**: Beginning of upward trend
- **2019-2024**: Higher plateau around 5.50-5.58

## Country-Level Analysis

### Countries with Largest Changes in Happiness

```{r country-changes}
most_changed_countries <- 
  whr |>
  drop_na(year, ladder_score) |> 
  group_by(country_name) |>
  summarise(
    first_year = min(year),
    last_year = max(year),
    first_score = first(ladder_score, order_by = year),
    last_score = last(ladder_score, order_by = year),
    change = last_score - first_score,
    years_span = last_year - first_year,
    .groups = 'drop'
  ) |>
  filter(years_span >= 10) |> 
  group_by(improvement = change > 0) |> 
  slice_max(order_by = abs(change), n = 5) |> 
  ungroup() 

improvements_plot <- 
  whr |>
  inner_join(
    most_changed_countries |> filter(improvement == TRUE), 
    by = join_by(country_name)
    ) |> 
  ggplot(aes(x = year, y = ladder_score, color = country_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_x_continuous(breaks = scales::breaks_width(3)) +
  labs(
    title = "Greatest improvements",
    x = "Year",
    y = "Happiness Score",
    color = NULL
  ) 

declines_plot <- 
  whr |>
  inner_join(
    most_changed_countries |> filter(improvement == FALSE), 
    by = join_by(country_name)
  ) |> 
  ggplot(aes(x = year, y = ladder_score, color = country_name)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_x_continuous(breaks = scales::breaks_width(3)) +
  scale_color_manual(
    values = 
        c("#d62728", "#ff7f0e", "#8c564b", "#e377c2", "#7f7f7f")
  ) +
  labs(
    title = "Greatest Declines",
    x = "Year", 
    y = "Happiness Score",
    color = NULL
  ) 


improvements_plot + 
    declines_plot & 
    theme_minimal() &
    theme(
      plot.title = element_text(size = 14, face = "bold"),
      axis.text = element_text(size = 11)
    ) 
```

#### Key findings

- Post-Soviet and eastern European countries showed improvement trends, with Serbia, Bulgaria, and Georgia having the largest improvements in happiness.
- Afghanistan, Lebanon, and Jordan had the most severe declines in happiness, coinciding with major political and economic events. 

## Happiness Contributing Factors

### Contributing factors 2024

For the years with complete factor data, we can analyze what drives happiness differences between countries.

```{r factor-analysis}
factor_contributions <- 
  whr |> 
  filter(year == 2024) |> 
  summarise(
    GDP = mean(explained_by_log_gdp_per_capita, na.rm = TRUE),
    `Social Support` = mean(explained_by_social_support, na.rm = TRUE),
    `Life Expectancy` = mean(explained_by_healthy_life_expectancy, na.rm = TRUE),
    `Freedom` = mean(explained_by_freedom_to_make_life_choices, na.rm = TRUE),
    `Generosity` = mean(explained_by_generosity, na.rm = TRUE),
    `Low Corruption` = mean(explained_by_perceptions_of_corruption, na.rm = TRUE)
  ) |>
  pivot_longer(
    everything(), 
    names_to = "Factor", 
    values_to = "Average_Contribution"
  ) |>
  arrange(desc(Average_Contribution))

factor_contributions |> 
ggplot(aes(x = reorder(Factor, Average_Contribution), y = Average_Contribution)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  coord_flip() +
  labs(
    title = "Average Contribution of Factors to Happiness",
    subtitle = "Based on World Happiness Report data (2019-2024)",
    x = "Factor",
    y = "Average Contribution to Happiness Score"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.text = element_text(size = 11)
  ) 
```

The top contributors to happiness worldwide in 2024 were social support, GDP per capita, and having a sense of freedom. 

### Happiness Factor Breakdown for Top 10 Countries (2024)

```{r top-countries}
# Get the top 10 happiest countries from 2024 and their factor breakdowns
top_countries_2024 <- 
  whr |>
  filter(year == 2024) |>
  slice_max(order_by = ladder_score, n = 10)

stacked_data <- 
  top_countries_2024 |>
  select(-ladder_score, -rank, -year, -upperwhisker, -lowerwhisker) |>
  pivot_longer(cols = -country_name, names_to = "Factor", values_to = "Contribution") |>
  mutate(
    Factor = case_when(
      Factor == "explained_by_log_gdp_per_capita" ~ "GDP per Capita",
      Factor == "explained_by_social_support" ~ "Social Support", 
      Factor == "explained_by_healthy_life_expectancy" ~ "Life Expectancy",
      Factor == "explained_by_freedom_to_make_life_choices" ~ "Freedom",
      Factor == "explained_by_generosity" ~ "Generosity",
      Factor == "explained_by_perceptions_of_corruption" ~ "Low Corruption",
      Factor == "dystopia_residual" ~ "Dystopia + residual",
      TRUE ~ Factor
    ) |> 
      fct_reorder(Contribution),
    country_name = 
        fct_reorder(country_name, Contribution, .fun = sum, .desc = TRUE)
  )

stacked_data |> 
  ggplot(aes(x = country_name, y = Contribution, fill = Factor)) +
  geom_col() +
  labs(
    title = "Happiness Factor Breakdown for Top 10 Happiest Countries",
    subtitle = "2024 World Happiness Report Data",
    x = "Country",
    y = "Contribution to Happiness Score",
    fill = "Contributing Factor"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "right"
  ) +
  scale_fill_brewer(type = "qual", palette = "Set2")
```


#### Patterns Among Top Countries

- **Nordic countries tend to be happy**: Finland (`{r} round(top_countries_2024$ladder_score[1], 2)`), Denmark (`{r} round(top_countries_2024$ladder_score[2], 2)`), Iceland (`{r} round(top_countries_2024$ladder_score[3], 2)`), Sweden (`{r} round(top_countries_2024$ladder_score[4], 2)`), and Norway (`{r} round(top_countries_2024$ladder_score[7], 2)`) show similar balanced patterns.
- **Effect of GDP varies**: Luxembourg has the highest GDP contribution, while Costa Rica has high happiness with lower GDP contribution, but strong social factors.
- **Balanced Approach**: No single happiness factor dominates. 

## Conclusions

1. **Global happiness is gradually improving**: The world has become slightly happier over the past decade.

2. **Social connections and economic conditions are most important**: GDP per capita and social support are the strongest contributors to national happiness. 

3. **Multiple pathways to happiness**: The top countries show that there are different combinations of factors that can lead to high happiness levels.

5. **Crises can severely affect happiness**: Countries experiencing political upheaval or economic crisis show dramatic happiness declines.

